---
title: Create awesome pivot table with R
author: ~
date: '2020-10-08'
slug: create-awesome-pivot-table
categories: [R]
tags: [R, pivot table, tidyquant]
---

# 1. Libraries

- `tidyverse` A collection of lots of tidy packages from Rstudio, e.g. `ggplot2`, `dplyr`, `stringr`, `purrr`, `forcat` ...

- `tidyquant` A package integrated the quantitative analysis functionality of `zoo`, `xts`, `quantmod`, `TTR`, and `PerformanceAnalytics`, in a tidy way.

- `gt` A package for creating attractive tables.


```{r load-libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(tidyquant)
library(kableExtra)
library(gt)
```

You can check the functions in a package by `ls("package:pkgname")`. For example.
```{r list-functions}
ls("package:tidyquant") %>% head()
```

# 2. Demo raw data

Stock data are retrieved here for demonstration with `tq-get` from `tidyquant` package, *AAPL*, *GOOG*, *NFLX*, *NVDA*.

```{r raw-data}
stock_raw <- c("AAPL", "GOOG", "NFLX", "NVDA") %>% 
  tq_get(from = "2015-01-01", to = "2020-10-01") %>% 
  select(symbol, date, adjusted)

```

Show the first and the last 50 records.

```{r}
stock_raw %>% 
  slice(1:50, (nrow(.)-49):nrow(.)) %>% 
  knitr::kable() %>% 
  kable_paper() %>%
  scroll_box(width = "100%", height = "400px")
```


# 3. Pivot table

Here create 3 different pivot tables to show the percent change by year for the given stocks.

## 3.1 Row: stockname, month; Column: year.
```{r row-stock-month-col-year}
stock_raw  %>%
    pivot_table(
        .rows    = c(~ symbol, ~ MONTH(date, label = TRUE)),
        .columns = ~ YEAR(date),
        .values  = ~ MEDIAN(adjusted)
    ) %>%
    rename_at(.vars = 1:2, ~ c("Symbol", "Month")) %>% 
  knitr::kable() %>% 
  kable_paper() %>%
  scroll_box(width = "100%", height = "400px")

```

## 3.2 Row: year, month; Column: stockname.

```{r row-year-month-col-stock}
stock_raw  %>%
    pivot_table(
        .rows    = c(~ YEAR(date), ~ MONTH(date, label = TRUE)),
        .columns = ~ symbol,
        .values  = ~ MEDIAN(adjusted)
    ) %>%
    rename_at(.vars = 1:2, ~ c("Year", "Month")) %>% 
  knitr::kable() %>% 
  kable_paper() %>%
  scroll_box(width = "100%", height = "400px")
```


## 3.3 Row: year, month; Column: stockname.

```{r row-stock-year-col-month}
stock_raw  %>%
    pivot_table(
        .rows    = c(~ symbol, ~ YEAR(date)),
        .columns =  ~ MONTH(date, label = TRUE),
        .values  = ~ MEDIAN(adjusted)
    ) %>%
    rename_at(.vars = 1:2, ~ c("Symbol","Year")) %>% 
  knitr::kable() %>% 
  kable_paper() %>%
  scroll_box(width = "100%", height = "400px")
```
## 3.4 Row: year; Column: stockname
```{r row-year-col-stock}
stock_raw  %>%
    pivot_table(
        .rows    = ~ YEAR(date),
        .columns =  ~ symbol,
        .values  = ~ PCT_CHANGE_FIRSTLAST(adjusted)
    ) %>%
    rename_at(.vars = 1, ~ c("Year")) %>% 
  knitr::kable() %>% 
  kable_paper() %>%
  scroll_box(width = "100%")

```

## 3.5 Row: stockname; Column:year 
```{r row-stock-col-year}
stock_raw  %>%
    pivot_table(
        .rows    = ~ symbol,
        .columns =  ~ YEAR(date),
        .values  = ~ PCT_CHANGE_FIRSTLAST(adjusted)
    ) %>%
    rename_at(.vars = 1, ~ c("Symbol")) %>% 
  knitr::kable() %>% 
  kable_paper() %>%
  scroll_box(width = "100%")

```

# Pivot chart
```{r pivot-chart}
color_fill <- "#1ecbe1"

pivot_table_gt <- stock_raw  %>%
    pivot_table(
        .rows    = ~ YEAR(date),
        .columns =  ~ symbol,
        .values  = ~ PCT_CHANGE_FIRSTLAST(adjusted)
    ) %>%
    rename_at(.vars = 1, ~ c("Year")) %>% 
    gt() %>%
    tab_header("Stock Returns", subtitle = md("_Technology Portfolio_")) %>%
    fmt_percent(columns = vars(AAPL, GOOG, NFLX, NVDA)) %>%
    tab_spanner(
        label = "Performance",
        columns = vars(AAPL, GOOG, NFLX, NVDA)
    ) %>%
    tab_source_note(
        source_note = md("_Data Source:_ Stock data retreived from Yahoo! Finance via tidyquant.")
    ) %>%
    tab_style(
        style = list(
            cell_fill(color = color_fill),
            cell_text(weight = "bold", color = "white")
        ),
        locations = cells_body(
            columns = vars(AAPL),
            rows    = AAPL >= 0)
    ) %>%
    tab_style(
        style = list(
            cell_fill(color = color_fill),
            cell_text(weight = "bold", color = "white")
        ),
        locations = cells_body(
            columns = vars(GOOG),
            rows    = GOOG >= 0)
    ) %>%
    tab_style(
        style = list(
            cell_fill(color = color_fill),
            cell_text(weight = "bold", color = "white")
        ),
        locations = cells_body(
            columns = vars(NFLX),
            rows    = NFLX >= 0)
    ) %>%
    tab_style(
        style = list(
            cell_fill(color = color_fill),
            cell_text(weight = "bold", color = "white")
        ),
        locations = cells_body(
            columns = vars(NVDA),
            rows    = NVDA >= 0)
    )

pivot_table_gt
```

You can save the chart with `gtsave`, maybe `Phantom JS` required.
```{r save-chart, eval=FALSE}
library(webshot)
pivot_table_gt %>% gtsave(filename = "folder/file.png")
```


# Reference
1. [How to make excel pivot tables in R](https://www.youtube.com/watch?v=K5qR-EREf_g)

1. [Create Awesome HTML Table with knitr::kable and kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)